Finding a closest point in a lattice of Voronoi's first kind

### 2014-07-22 reviews/response.tex

``A final remark is that our algorithm assumes that the obtuse superbasis is known in advance.  It is known that all lattices of dimension less than or equal to 3 are of Voronoi's first kind and an algorithm exists to recover the obtuse superbasis in this case~\cite{SPLAG}.  Lattices of dimension larger than 3 need not be of Voronoi's first kind.  Given a lattice, is it possible to efficiently decide whether it is of Voronoi's first kind?  Is it possible to efficiently find an obtuse superbasis if it exists?  It is suspected that the answer to this second question is no because an efficient solution would yield a solution to a known problem, that of determining whether a lattice is rectangular (has a basis consisting of pairwise orthogonal vectors) given an arbitrary basis~\cite{Lenstra_Silverberg_revisting_gentra_szydlo_2014}.''  %It is suspected that this simpler problem is difficult.''

\item\textbf{Comment:} 
This paper makes a solid contribution to the theory of lattice problems. I think the results here are
well-motivated (I encourage the authors to include some of the motivation mentioned in the summary), and point to the possibility that there might be large and useful classes of lattices where
solving CVP is “easy”.
\\\textbf{Response:}
The following paragraph has been added to Section~\zref{sec:discussion} to suggest that similar fast algorithms for other classes of lattices might exist:

``Another interesting question is: are there subfamilies of Voronoi's first kind that admit even faster algorithms?  Both $A_n$ and $A_n^*$ are examples of this, but there might exist other subfamilies with algorithms faster than $O(n^4)$.  A related question is: can the techniques developed in this paper be applied to other families of lattices, i.e., beyond just those of Voronoi's first kind?''

\item\textbf{Comment:} 
On a technical level, I found the main convergence proof to be interesting, as I've rarely seen
combinatorial arguments successfully applied to lattice problems. On the other hand, while I've
convinced myself the authors have the ``right'' convergence proof, I found its presentation to be
rather awkward and unintuitive. I highly recommend that the authors rephrase the proof using
the language of Laplacians, which would make things more clear (see the additional comments
below for some technical suggestions).
\\\textbf{Response:}
We thank the reviewer for their comments.  The order of some paragraphs and lemmas has been changed in Section~\zref{sec:seri-relev-vect} with the intent of improving the readability of the proofs and presenting the intuition behind the proofs before formalities begin.  We agree that a simpler proof would always be preferable, but, after considerable effort, have not found a simpler rigorous proof.  We respond to the reviewers numbered comments below.

@@

\end{itemize}

\begin{enumerate}

\item\textbf{Comment:} 
I believe Proposition 2.1 and 2.2 appear essentially verbatim in [8]. Same for Propositions
3.1 (which is essentially by definition) and 3.2, they appear in [9]. I don’t think it adds any
readability/anything of value to reprove them here.
\\\textbf{Response:}

We agree that these propositions are well known/straightforward, but would prefer to keep them so that the paper is self contained. We do not know of references to specific proofs that we can give for these (unlike Theorem~4.1 for which a proof is directly given in \cite[Theorem~3]{ConwaySloane1992_voronoi_lattice_3d_obtuse_superbases}).  The proof of Proposition 2.1 has been shortened in response to comment~\ref{com:rev2halfopen} of reviewer 2 and the proof of Proposition 2.2 has also been shortened.  A result like Proposition 2.2 does appear informally in~\cite[Sec~5]{Shalvi_iterativeslicer_2009}.  As far as we are aware, the remarkable algorithm in~\cite{MicciancioVoulgaris_deterministic_jv_2013} does not rely upon bounding the number of lattice points inside a sphere (or any other volume).  Instead, the procedure operates recursively on scaled versions of the lattice.  The scaling is chosen to ensure that the target point always lies inside twice the Voronoi cell.

@@

\item\textbf{Comment:} \label{com:provethm4.1}
Furthermore, one can use this to give a two line proof of Theorem 4.1.  This theorem I actually do think make sense to reprove here, because modifications of this proof are what drives the convergence bound. 
\\\textbf{Response:}

The property described by the reviewer is indeed needed to give a proof of Theorem 4.1.  To our knowledge, one also needs the property that the relevant vectors of a lattice are given by the shortest coset representatives of $\Lambda / 2\Lambda$.  This result is attributed to Voronoi (see for example \cite[Theorem~2]{ConwaySloane1992_voronoi_lattice_3d_obtuse_superbases} and \cite[Theorem~3.2]{MicciancioVoulgaris_deterministic_jv_2013}).  This is used by Conway and Sloane in~\cite[Theorem~3]{ConwaySloane1992_voronoi_lattice_3d_obtuse_superbases}.

We believe that the proof, whilst not difficult, is ultimately a distraction from our main results.  This would be different if properties needed for the proof, such as the function $\subrng_1$ and Selling's formula were also needed for proving the convergence of our algorithm.  However, as we have stated in comment~\ref{com:uselaplaicesubr1} above, Selling's formula is not directly needed, and $\subrng_2$, not $\subrng_1$ is the function required.

%The order of some paragraphs and lemmas has been changed in Section~\zref{sec:seri-relev-vect} with the intent of improving the readability of the proofs and presenting the intuition behind the proofs before formalities begin.

@@

We believe that it is only Theorem 3 from~\cite{ConwaySloane1992_voronoi_lattice_3d_obtuse_superbases} that needs to cited.  Theorem~2 in \cite{ConwaySloane1992_voronoi_lattice_3d_obtuse_superbases} asserts that the relevant vectors in any lattice $\Lambda$ are the shortest coset representatives of the quotient $\Lambda/2\Lambda$.  Theorem~3 uses this, Selling's formula, and the fact that the Selling parameters are nonpositive to prove that the relevant vectors in a lattice of Voronoi's first kind take the form
\[
\sum_{i \in I} b_i
\]
where $I$ is a strict subset of $\{1, 2, \dots, n+1\}$ that is not empty.

%BLERG: Check that Theorems 2 and 3 in the published version of\cite{ConwaySloane1992_voronoi_lattice_3d_obtuse_superbases} are what I think they are!

\item\textbf{Comment:}  
page 8, lemma 5.1. I think you forgot a factor 2 preceding the double summations over S and T (at end of lemma statement, and also in the proof.) 
\\\textbf{Response:}
This error has been fixed.

\item\textbf{Comment:}  
page 11, ``with be'' $\to$ ``will be'' 
\\\textbf{Response:}
Fixed.

\item\textbf{Comment:}  
Displayed equation before ``Putting p=...'', A $|$ should be $\|$
\\\textbf{Response:}
Fixed.

\item\textbf{Comment:}  
(5.7) again, missing factor 2 before double summation, and second set should be T, not S. 
\\\textbf{Response:}
Fixed.

### 2014-07-22 paper.tex
minor

\begin{lemma}\label{lem:rngdecreases}
Let $Bu$ with $u \in \ints^{n+1}$ be a lattice point that is $\ell$-close to $y = Bz$ where $\ell > 0$.  Let $g \in \{0,1\}^{n+1}$ be such that
\begin{equation}\label{eq:qismin}
\|B(z - u - g)\|^2 = \min_{t \in \{0,1\}^{n+1}}\|B(z - u - t)\|^2.
\end{equation}
The lattice point $B(u+g)$ is $(\ell-1)$-close to $y$.
\end{lemma}
\begin{proof}
Because $Bu$ is $\ell$-close to $y$ there exists $v \in \ints^{n+1}$ with $\rng(v) = \ell$ such that $B(u+v)$ is a closest lattice point to $y$.  Define subsets of indices
\[
S = \{i \mid g_i = 0, v_i = \max(v) \}, \qquad T = \{i \mid g_i = 1, v_i = \min(v) \}, 
\]
and put $h = g + \onebf_S - \onebf_T \in \{0,1\}^{n+1}$ and $w = v - h$.  Observe that $h_i = 0$ whenever $v_i = \min(v)$ and so $\min(w) = \min(v-h) = \min(v)$.  Also, $h_i = 1$ whenever $v_i = \max(v)$ and so $\max(w) = \max(v-h) = \max(v) - 1$.  Thus,
\[
\rng(w) = \max(v) - 1 - \min(v) = \rng(v) - 1 = \ell - 1.
\]
The lemma will follow if we show that $B(u+g+w)$ is a closest lattice point to $y$ since then the lattice point $B(u+g)$ with be $(\ell-1)$-close to $y$.  The proof is by contradiction.  Suppose $B(u+g+w)$ is not a closest point to $y$, that is, suppose
\[
\|B(z-u-g-w)\|^2 > \|B(z-u-v)\|^2.
\]
Putting $p = z-u-v$ we have
\[
\|B(p+\onebf_S-\onebf_T)\|^2 > \|Bp\|^2.
\]
By part~\ref{eq:lem:decSellingsincdec} of Lemma~\ref{lem:decSellings} we obtain
\begin{equation}\label{eq:BpBpdecincineq}
\|Bp\|^2 - \|B(p+\onebf_S-\onebf_T)\|^2 = \Phi(S,p) + \Phi(\bar{T}, p) + 2\sum_{i\in S}\sum_{j \in T} q_{ij} < 0.
\end{equation}

@@

We now show how $W(C,\bar{C})$ can be represented as a binary quadratic form.  Put $t_0 = 1$ and $t_{n+2} = 0$ and
\[
t_i = \begin{cases}
1, & i \in C \\
0, & i \in \bar{C}
\end{cases}
\]
for $i = 1,2,\dots,n+1$.  Observe that
\[
t_i(1 - t_j) = \begin{cases}
1, & i \in C, j \in \bar{C} \\
0, & \text{otherwise}.
\end{cases}
\]
The weight can now be written as
\begin{align*}
W(C,\bar{C}) = \sum_{i \in C} \sum_{j \in \bar{C}} w_{ij} = \sum_{i =0}^{n+2} \sum_{j =0}^{n+2} w_{ij} t_i (1 - t_j) = F(t),
\end{align*}
say.  Finding a minimum cut is equivalent to finding the binary vector $t = (t_1, \dots, t_{n+1})$ that minimises $F(t)$.  Write,
\[
F(t) =  \sum_{i=0}^{n+2} \sum_{j =0}^{n+2} w_{ij}t_i - \sum_{i=0}^{n+2} \sum_{j =0}^{n+2} w_{ij} t_it_j.
\]
Letting $k_i = \sum_{j =0}^{n+2} w_{ij}$, and using that $t_0 = 1$ and $t_{n+2} = 0$,
\[
F(t) = \sum_{i=0}^{n+1}k_it_i  - w_{00} - \sum_{i=1}^{n+1} w_{i0} t_i - \sum_{j=1}^{n+1} w_{0j} t_j - \sum_{i=1}^{n+1} \sum_{j =1}^{n+1} w_{ij} t_it_j.
\]
Because $w_{00} = 0$ and $w_{ij} = w_{ji}$ we have
\[
F(t) = k_0 + \sum_{i=1}^{n+1}( k_i  - 2 w_{i0}) t_i  - \sum_{i=1}^{n+1} \sum_{j =1}^{n+1} w_{ij} t_it_j.
\]
The constant term $k_0$ is unimportant for the purpose of minimisation so finding a minimum cut is equivalent to minimising the binary quadratic form
\[
\sum_{i=1}^{n+1}g_i t_i  - \sum_{i=1}^{n+1} \sum_{j =1}^{n+1} w_{ij} t_it_j,
\]
where $g_i = k_i  - 2 w_{i0} = d_i + \sum_{j=1}^{n+1} w_{ij}$.  It only remains to observe the equivalence of this quadratic form and $Q(t)$ from~\eqref{eq:quadformnp} when the weights are assigned to satisfy,
\begin{align*}
&q_{ij} = - w_{ij} \qquad i,j = 1,\dots,n+1 \\
&s_i = g_i = d_i + \sum_{j=1}^{n+1} w_{ij}.
\end{align*}
Because the $q_{ij}$ are nonpositive for $i \neq j$ the weights $w_{ij}$ are nonnegative for all $i \neq j$ with $i,j = 1,\dots,n+1$.  As discussed the value of the weights $w_{ii}$ have no effect on the weight of any cut $W$ so setting $q_{ii} = - w_{ii}$ for  $i = 1,\dots,n+1$ is of no consequence.  Finally the weights $w_{i,n+2}$ and $w_{0,i}$ can be chosen so that both are nonnegative and 
\[
w_{i,n+2} - w_{0,i} = d_i = s_i + \sum_{j=1}^{n+1} q_{ij} = s_i
\]  
because $\sum_{j=1}^{n+1} q_{ij} = 0$ due to the superbase condition~\eqref{eq:superbasecond}.  That is, we choose $w_{i,n+2} = s_i$ and $w_{0,i} = 0$ when $s_i \geq 0$ and $w_{i,n+2}=0$ and $w_{0,i} = -s_i$ when $s_i < 0$.  With these choices, all the weights $w_{ij}$ for $i \neq j$ are nonnegative.  A minimiser of $Q(t)$, and correspondingly a solution of~\eqref{eq:pvecmin} can be computed in $O(n^3)$ operations by computing a minimum cut in the undirected flow network $G$ assigned with these nonnegative weights~\cite{Picard_min_cuts_1974,Sankaran_solving_CDMA_mincut_1998,Ulukus_cdma_mincut_1998,Cormen2001}.  %The closest lattice point is then computed using~\eqref{eq:compnp}.

>>>>

\begin{lemma}\label{lem:rngdecreases}
Let $Bu$ with $u \in \ints^{n+1}$ be a lattice point that is $\ell$-close to $y = Bz$ where $\ell > 0$.  Let $g \in \{0,1\}^{n+1}$ be such that
\begin{equation}\label{eq:qismin}
\|B(z - u - g)\|^2 = \min_{t \in \{0,1\}^{n+1}}\|B(z - u - t)\|^2.
\end{equation}
The lattice point $B(u+g)$ is $(\ell-1)$-close to $y$.
\end{lemma}
\begin{proof}
Because $Bu$ is $\ell$-close to $y$ there exists $v \in \ints^{n+1}$ with $\rng(v) = \ell$ such that $B(u+v)$ is a closest lattice point to $y$.  Define subsets of indices
\[
S = \{i \mid g_i = 0, v_i = \max(v) \}, \qquad T = \{i \mid g_i = 1, v_i = \min(v) \}, 
\]
and put $h = g + \onebf_S - \onebf_T \in \{0,1\}^{n+1}$ and $w = v - h$.  Observe that $h_i = 0$ whenever $v_i = \min(v)$ and so $\min(w) = \min(v-h) = \min(v)$.  Also, $h_i = 1$ whenever $v_i = \max(v)$ and so $\max(w) = \max(v-h) = \max(v) - 1$.  Thus,
\[
\rng(w) = \max(v) - 1 - \min(v) = \rng(v) - 1 = \ell - 1.
\]
The lemma will follow if we show that $B(u+g+w)$ is a closest lattice point to $y$ since then the lattice point $B(u+g)$ will be $(\ell-1)$-close to $y$.  The proof is by contradiction.  Suppose $B(u+g+w)$ is not a closest point to $y$, that is, suppose
\[
\|B(z-u-g-w)\|^2 > \|B(z-u-v)\|^2.
\]
Putting $p = z-u-v$ we have
\[
\|B(p+\onebf_S-\onebf_T)\|^2 > \|Bp\|^2.
\]
By part~\ref{eq:lem:decSellingsincdec} of Lemma~\ref{lem:decSellings} we obtain
\begin{equation}\label{eq:BpBpdecincineq}
\|Bp\|^2 - \|B(p+\onebf_S-\onebf_T)\|^2 = \Phi(S,p) + \Phi(\bar{T}, p) + 2\sum_{i\in S}\sum_{j \in T} q_{ij} < 0.
\end{equation}

@@

We now show how $W(C,\bar{C})$ can be represented as a binary quadratic form.  Put $t_0 = 1$ and $t_{n+2} = 0$ and
\[
t_i = \begin{cases}
1, & i \in C \\
0, & i \in \bar{C}
\end{cases}
\]
for $i = 1,2,\dots,n+1$.  Observe that
\[
t_i(1 - t_j) = \begin{cases}
1, & i \in C, j \in \bar{C} \\
0, & \text{otherwise}.
\end{cases}
\]
The weight can now be written as
\begin{align*}
W(C,\bar{C}) = \sum_{i \in C} \sum_{j \in \bar{C}} w_{ij} = \sum_{i =0}^{n+2} \sum_{j =0}^{n+2} w_{ij} t_i (1 - t_j) = F(t),
\end{align*}
say.  Finding a minimum cut is equivalent to finding the binary vector $t = (t_1, \dots, t_{n+1})$ that minimises $F(t)$.  Write,
\[
F(t) =  \sum_{i=0}^{n+2} \sum_{j =0}^{n+2} w_{ij}t_i - \sum_{i=0}^{n+2} \sum_{j =0}^{n+2} w_{ij} t_it_j.
\]
Letting $k_i = \sum_{j =0}^{n+2} w_{ij}$, and using that $t_0 = 1$ and $t_{n+2} = 0$,
\[
F(t) = \sum_{i=0}^{n+1}k_it_i  - w_{00} - \sum_{i=1}^{n+1} w_{i0} t_i - \sum_{j=1}^{n+1} w_{0j} t_j - \sum_{i=1}^{n+1} \sum_{j =1}^{n+1} w_{ij} t_it_j.
\]
Because $w_{00} = 0$ and $w_{ij} = w_{ji}$ we have
\[
F(t) = k_0 + \sum_{i=1}^{n+1}( k_i  - 2 w_{i0}) t_i  - \sum_{i=1}^{n+1} \sum_{j =1}^{n+1} w_{ij} t_it_j.
\]
The constant term $k_0$ is unimportant for the purpose of minimisation so finding a minimum cut is equivalent to minimising the binary quadratic form
\[
\sum_{i=1}^{n+1}g_i t_i  - \sum_{i=1}^{n+1} \sum_{j =1}^{n+1} w_{ij} t_it_j,
\]
where $g_i = k_i  - 2 w_{i0} = d_i + \sum_{j=1}^{n+1} w_{ij}$.  It only remains to observe the equivalence of this quadratic form and $Q(t)$ from~\eqref{eq:quadformnp} when the weights are assigned to satisfy,
\begin{align*}
&q_{ij} = - w_{ij} \qquad i,j = 1,\dots,n+1 \\
&s_i = g_i = d_i + \sum_{j=1}^{n+1} w_{ij}.
\end{align*}
Because the $q_{ij}$ are nonpositive for $i \neq j$ the weights $w_{ij}$ are nonnegative for all $i \neq j$ with $i,j = 1,\dots,n+1$.  As discussed the value of the weights $w_{ii}$ has no effect on the weight of any cut $W$ so setting $q_{ii} = - w_{ii}$ for  $i = 1,\dots,n+1$ is of no consequence.  Finally the weights $w_{i,n+2}$ and $w_{0,i}$ can be chosen so that both are nonnegative and 
\[
w_{i,n+2} - w_{0,i} = d_i = s_i + \sum_{j=1}^{n+1} q_{ij} = s_i
\]  
because $\sum_{j=1}^{n+1} q_{ij} = 0$ due to the superbase condition~\eqref{eq:superbasecond}.  That is, we choose $w_{i,n+2} = s_i$ and $w_{0,i} = 0$ when $s_i \geq 0$ and $w_{i,n+2}=0$ and $w_{0,i} = -s_i$ when $s_i < 0$.  With these choices, all the weights $w_{ij}$ for $i \neq j$ are nonnegative.  A minimiser of $Q(t)$, and correspondingly a solution of~\eqref{eq:pvecmin} can be computed in $O(n^3)$ operations by computing a minimum cut in the undirected flow network $G$ assigned with these nonnegative weights~\cite{Picard_min_cuts_1974,Sankaran_solving_CDMA_mincut_1998,Ulukus_cdma_mincut_1998,Cormen2001}.  %The closest lattice point is then computed using~\eqref{eq:compnp}.
